Supplementary Material

### 2021-10-22 analysis/paper/paper.Rmd
correct SD miscalculation

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!-- with Z-score standardisation -->

Pearson's *r* suggests a 
`r strength_cor` `r direct_cor` 
correlation between the total weight of the biofilms and the total starch count 
(standardised by z-score) extracted from the samples across treatments, 
*r* = `r cor_r`, 
90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
p `r cor_p`
(Figure \@ref(fig:cor-plot)).

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```

The same test was applied to total biofilm weight and starch count per mg 
calculus (also standardised by z-score), resulting in a weak positive correlation, 
*r* = `r cor2_r`, 
90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
p `r cor2_p`
(Figure \@ref(fig:cor-plot2)).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2)).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.

>>>>

```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

Pearson's *r* suggests a 
`r strength_cor` `r direct_cor` 
correlation between the total weight of the biofilms and the total starch count 
(standardised by z-score) extracted from the samples across treatments, 
*r* = `r cor_r`, 
90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
p `r cor_p`
(Figure \@ref(fig:cor-plot)A).

<!--
```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```
-->

The same test was applied to total biofilm weight and starch count per mg 
calculus (also standardised by z-score), resulting in a weak positive correlation, 
*r* = `r cor2_r`, 
90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
p `r cor2_p`
(Figure \@ref(fig:cor-plot)B).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.

### 2021-10-22 analysis/paper/paper.tex
remove redundant code from paper and R scripts
correct SD miscalculation

It was not possible to differentiate between potato and
wheat starches smaller than ca. 10 \(\mu\)m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the total amount of small starches in these two solutions,
0.653\%
are from wheat.

@@

\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.}
\centering
\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
\hline
Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
\hline
mix & potato &  &  & 1959 (79.6\%) & 1801 & 501 (20.40\%) & 446 & 2460 (100\%) & 2189\\
\hline
mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6026 & 1381 (7.93\%) & 1196 & 17417 (100\%) & 15878\\
\hline
mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 4929 & 1882 (9.47\%) & 989 & 19877 (100\%) & 13451\\
\hline
potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2402 & 930 (19.20\%) & 929 & 4846 (100\%) & 3316\\
\hline
wheat & wheat & 15235 (55.00\%) & 11944 & 12148 (43.9\%) & 11052 & 1953 (7.06\%) & 2016 & 27680 (100\%) & 23554\\
\hline
\end{tabular}
\end{table}

@@

\caption{Scatter plot of sample weight and standardised starch count by z-score for separated treatments.}\label{fig:cor-plot}
\end{figure}

Pearson's \emph{r} suggests a
strong positive
correlation between the total weight of the biofilms and the total starch count
(standardised by z-score) extracted from the samples across treatments,
\emph{r} = 0.659,
90\%CI{[}0.463, 0.794{]},
p \textless{} 0.001
(Figure \ref{fig:cor-plot}).

{\centering \includegraphics{../figures/cor-plot2-1}

\caption{Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot2}
\end{figure}

The same test was applied to total biofilm weight and starch count per mg
calculus (also standardised by z-score), resulting in a weak positive correlation,
\emph{r} = 0.3,
90\%CI{[}0.0618, 0.506{]},
p = 0.0403
(Figure \ref{fig:cor-plot2}).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot});
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot2}).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

>>>>

It was not possible to differentiate between potato and
wheat starches smaller than ca. 10 \(\mu\)m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the total amount of small starches in these two solutions,
99.2\%
are from wheat.

@@

\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.}
\centering
\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
\hline
Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
\hline
mix & potato &  &  & 1959 (79.6\%) & 1801 & 501 (20.40\%) & 446 & 2460 (100\%) & 2189\\
\hline
mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6026 & 1381 (7.93\%) & 1196 & 17417 (100\%) & 15878\\
\hline
mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7653 & 1882 (9.47\%) & 1596 & 19877 (100\%) & 17768\\
\hline
potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2402 & 930 (19.20\%) & 929 & 4846 (100\%) & 3316\\
\hline
wheat & wheat & 15235 (55.00\%) & 11944 & 12148 (43.9\%) & 11052 & 1953 (7.06\%) & 2016 & 27680 (100\%) & 23554\\
\hline
\end{tabular}
\end{table}

@@

\caption{Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot}
\end{figure}

Pearson's \emph{r} suggests a
strong positive
correlation between the total weight of the biofilms and the total starch count
(standardised by z-score) extracted from the samples across treatments,
\emph{r} = 0.659,
90\%CI{[}0.463, 0.794{]},
p \textless{} 0.001
(Figure \ref{fig:cor-plot}A).

The same test was applied to total biofilm weight and starch count per mg
calculus (also standardised by z-score), resulting in a weak positive correlation,
\emph{r} = 0.3,
90\%CI{[}0.0618, 0.506{]},
p = 0.0403
(Figure \ref{fig:cor-plot}B).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot}A);
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

### 2021-10-22 analysis/supplementary-materials/supp-mat.Rmd
correct SD miscalculation

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(patchwork)
source(here("analysis/scripts/analysis.R"))
#raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
#raw_data_solutions <- read_csv(here("analysis/data/raw_data/solution_counts.csv"))
```

@@

Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 &deg;C.

@@

```{r amylase-activity-tables}
plt1_ph1_result
plt1_ph2_result
plt2_ph1_result
plt2_ph2_result
```

@@

```{r solution-bar-pl}
sol_comb_long %>%
  filter(size == "total") %>%
  group_by(treatment, starch) %>%
  ggplot(aes(x = treatment, y = count, fill = treatment, col = starch)) +
  geom_col(size = 1.5) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "grey"),
        panel.grid.major.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_fill_viridis_d() +
  scale_color_viridis_d(begin = 0.5)
```

@@

Size distribution within the samples:

```{r message=FALSE, warning=FALSE}
samp_size_pl <- corr_counts_long %>%
  filter(size != "total",
         treatment != "control") %>%
  group_by(treatment, starch, size) %>%
  summarise(count = mean(count, na.rm = T)) %>%
  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
  ggplot(aes(x = starch, y = percent, fill = size)) +
    geom_col(position = "dodge") +
    facet_wrap(~ treatment, scales = "free_x") +
    scale_fill_viridis_d() +
    theme_bw()
```

>>>>

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(patchwork)
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/figures.R"))
#raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
#raw_data_solutions <- read_csv(here("analysis/data/raw_data/solution_counts.csv"))
```

@@

Amylase activity in U/mL enzyme, where U is mg maltose released from starch in six minutes at 36 &deg;C.

@@

```{r amylase-activity-tables}
plt1_ph1_result # plate 1, photometric reading 1
plt1_ph2_result # plate 1, photometric reading 2
plt2_ph1_result # plate 2, photometric reading 1
plt2_ph2_result # plate 2, photometric reading 2
```

@@

```{r solution-bar-pl}
sol_long %>%
  filter(size == "total") %>%
  group_by(treatment, starch) %>%
  ggplot(aes(x = treatment, y = count, fill = treatment, col = starch)) +
  geom_col(size = 1.5) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "grey"),
        panel.grid.major.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_fill_viridis_d() +
  scale_color_viridis_d(begin = 0.5)
```

@@

Separated correlation plots:

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
pl_cor
```

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```
